HO unification from object language to meta language

### 2024-03-27 paper/main.tex
rework eta intro

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}\xspace}

@@

\begin{invariant}[Link left hand side]\ref{inv:linklhs}
  The left hand side of a new link
  is a variable.
\end{invariant}

% TODO: io sposteri l'invariante sopra nella maybe-eta, perchè  
% la proprietà è vera solo per i link eta
% per i link beta, questo non è vero se sospendiamo i link
\noindent
If the variable is assigned during a run the link is considered for
progress and possibly eliminated. This is discussed in
section~\ref{sec:eta}.\marginpar{is this the right section?}

@@

$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
$\eta$-expanded version represented by the term $t$. Both the meta and the
object language accept terms unification up to $\eta$ equivalence, but as
sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
able to capture this equivalence property when using terms in \Fo. The main
reason is that the unification of two terms in \Fo{} appears to be syntactical
on the constructors of these terms. As a very concise example, \Ue can't unify
the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
rigid constructor \elpiIn{lam} is different from the rigid constructor
\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
order to address this problem. In the sections below, we explain how we detect
and compile \maybeeta terms with the corresponding \linketa and how these links
are used to make unification progress. The same explanation structure will be 
used for \maybebeta terms.

@@

\begin{invariant}[\linketa \lhs]
  The \lhs of a \linketa is a variable.
  \label{invariant:link-eta-left}
\end{invariant}

TODO: motivate this?

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F \appsep x$
  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

TODO: motivate this?

@@

TODO: dire che scendiamo sotto i vari abs che formano il contesto?

We can note that the insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since, if the link is woken up,
the unification of the terms and the suppression of the link form \linkStore
are terminating operation. Note that if the link remain suspend
\elpiIn{progress} continues to terminate, since, it would mean that the
condition of the \elpiIn{if}-branchement succeds and the \elpiIn{progress}
terminates its execution.

A second way to progress \linketa is to check if \linkStore contains two
\linketa $l_1$ and $l_2$ with same \lhs. In this case we are allowed to unify

TODO: dire che termina

@@

\begin{textcode}
  La deduplicate eta:
  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2]
    (a destra non c'è mai un termine con testa rigida)
  - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
    NOTA!! A dx abbiamo sempre un termine della forma lam x.VAR x!!!
    Altrimenti il link sarebbe stato risolto!!
  - dopo l'unificazione rimane un link [variabile] -> [etaX]
  - nella progress-eta, se a sx abbiamo una constante o un'app, allora eta-espandiamo
    di uno per poter unificare con il termine di dx.
\end{textcode}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}

@@

\begin{invariant}[Link left hand side]\label{inv:linklhs}
  The left hand side of a new link
  is a variable.
\end{invariant}

\noindent
If the variable is assigned during a run the link is considered for
progress and possibly eliminated. This is discussed in
section~\ref{sec:eta}.\marginpar{is this the right section?}

@@

$\eta$-reduction is an equivalence relation over terms where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
version of $t$. The implementation of the \elpiIn{comp} relation given in
\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
unification problem by adapting the \elpiIn{comp} relation such that it
recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

\noindent
where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
(\lhs) is linked with its left counterpart (\rhs).

\linketa are added in the link store (\linkStore) and activated when special
conditions are satisfied on \lhs or \rhs. These link activations are managed by
extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
add the following invariant:

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F_x$
  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how \linketa are activated during the execution of the program
and provide justification for why these two invariants remain true.

\subsection{Detection of \maybeeta}

@@

@@

TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$

A second way to progress \linketa, that we call \linketa deduplication, is to
check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
situation occurs if two \maybeeta terms are unified with a same unification
variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
construction are both on the form $\lambda x. T_x$) and remove one of the two
links.

TODO: example for this: $\lambda x. \lambda y. F \appsep y \appsep x = X,
\lambda x. \lambda y. F \appsep y \appsep x = Y$

We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since, a link activation runs
terminating operations (such as term unification and link-removal); 2) \linketa
deduplication runs again terminating operations and 3) if none of these two 
situations is performed, then the substitution and \linkStore
Note that if the link remain suspended \elpiIn{progress} continues to terminate,
since, it would mean that the condition of the \elpiIn{if}-branchement succeds
and the \elpiIn{progress} terminates its execution.

@@
